Parsing Postfix log files

### 2007-03-23 doc/logparser.tex
Improved parser overview. Move rule characteristics to rules section. Split rule attributes into their own subsection, and seriously improved them. Added subsections for the flow chart and a detailed description of actions; the latter is empty.

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements, one
can implement one's own processing.

@@

Before getting into detail about the parser, a brief overview is in order.
The parser is split into two parts: the parsing algorithm, and the rules
which are applied to the lines.

    \item [Rules] The first matching rule wins: no further rules are tried
        against that line,\footnote{With one exception, detailed later.}
        but there is a facility for the specifying the order of the rules,
        so that more specific rules can be matched first.  Rules are
        completely self-contained: there are no sub-rules, so the
        complexity is linear.  Rules are similar to the tokeniser/Lex part
        of a Lex/YACC parser.

    \item [Algorithm]

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the algorithm source code (significantly lowering the barrier to entry
for new users), and greatly simplifies both algorithm and rules.  The
separation also allows for a clear separation of functionality: rules
handle low level details of identifying a line, whereas the algorithm
handles the higher level details of following the path a mail takes through
postfix, assembling the required information, etc.

Rules are relatively simple, defining the following for each rule:

@@

    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match, or worse, might match unintentionally.

    \item [regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies writing
        rules and avoids needless repetition of complex regex components.

    \item [result\_cols and connection\_cols] This is how we extract the
        fields in the log line matched by the regex.  The format is:
        \newline 
        hostname = 1; helo = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the column
        name on the left and the match from the regex (\$1, \$2 etc) on the
        right hand side (without \$).

    \item [action] The action for the parser to take.

    \item [queueid] The index of the field in regex giving the queueid.

    \item [rule\_order] This is used to sort the rules, so that rules
        matching more commonly occurring log lines will be tried first, and
        is updated every time the program is run.

@@

    \item [result]  The result the log line represents: rejection,
        acceptance, info, etc.

    \item [result\_data and connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format is
        the same as result\_cols and connection\_cols.

@@

@@

Lets start with a flow chart describing the program:

>>>>

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleaned from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements, one can
implement one's own processing.

@@

Before getting into detail about the parser, a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the
tokeniser/Lex part of a Lex/YACC parser; they identify the line and return
some data, which the algorithm (the YACC part) receives, performing the
correct action.  Rules are all about identifying a line and extracting data
from it, whereas the algorithm is all about following the path the mail
takes through Postfix, piecing the data together into a coherent whole, and
performing housekeeping duties.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without modifications to the
algorithm source code (significantly lowering the barrier to entry for new
or casual users), and greatly simplifies both algorithm and rules.  It also
allows for a clear separation of functionality: rules handle low level
details of identifying a line, whereas the algorithm handles the higher
level details of following the path a mail takes through postfix,
assembling the required data, etc.

Rule characteristics:

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, detailed later in Section
        \ref{actions-in-detail}.} but there is a facility for the specifying
        the order of the rules, so that more specific rules can be matched
        first.

    \item Rules are completely self-contained: there are no sub-rules, so
        rules have linear complexity, and can be understood without
        reference to any other rules.

\subsection{Rule attributes}

Each rule defines the following:

@@

    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match the line, or worse, might match unintentionally.

    \item [regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies reading
        and writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.

    \item [result\_cols, connection\_cols] This specifies how the fields in
        the log line will be extracted.  The format is: \newline
        client\_hostname = 1; helo = 2; sender = 4; \newline
        i.e.  semi-colon separated assignment statements, with the variable
        name on the left and the match from the regex (\$1, \$2 etc.) on the
        right hand side (without \$).

    \item [result\_data, connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format of
        the same as result\_cols and connection\_cols.

    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
        Postfix would have taken to generate this line.  Info is represents
        an unspecified intermediate action that the parser is not
        interested in per se, but which does log useful information.  The
        extracted information will be used, but the other results will be
        discarded.

    \item [action] The action the algorithm will take, e.g.
        \begin{description}

            \item [ignore] Ignore this line.  The simplest action, it is
                required because every line from a program of interest to
                the parser must be parsed properly; any unparsed line is
                considered an error.

            \item [save\_by\_pid] Use the pid from the log line to lookup a
                connection, then save the data from the log line in the
                connection data structure.

            \item [commit] The mail is finished with, enter it in the
                database.

        A full list can be found in Section \ref{actions-in-detail}.

    \item [queueid] Specifies the match from the regex which gives the
        queueid, or zero if the log line doesn't contain a queueid.

    \item [rule\_order] This is an efficiency measure.  A counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in ascending rule\_order, and at the end of the run
        updates every rule's rule\_order so that rules which matched more
        frequently have a lower rule\_order.  Assuming that the 
        distribution of log lines is reasonably consistent, rules matching
        more commonly occurring log lines will be tried before rules
        matching less commonly occurring log lines, lowering the program's
        execution time.  XXX GIVE EXAMPLE TIMINGS XXX

@@

@@

\subsection{Algorithm flow chart}

@@

\label{full-algorithm}

\subsection{Actions in detail}

\label{actions-in-detail}
